---
title: "Parquet files"
teaching: 0
exercises: 0
---

:::::::::::::::::::::::::::::::::::::: questions 

- What is a parquet file?

- How to explore and open parquet files in R?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Understand the structure of parquet files.

- Learn how to read parquet files in R.


::::::::::::::::::::::::::::::::::::::::::::::::

## Introduction

In this episode, we will explore parquet files, a popular file format for storing large datasets efficiently. We will learn how to read parquet files in R and understand their structure.

::::::::::::::::::::::::::::::::::::::::::::::::callout

For this episode we will be using a sample OMOP CDM database that is pre-loaded with data. This database is a simplified version of a real-world OMOP CDM database and is intended for educational purposes only.

(UCLH only) This will come in a similar form as you would get data if you asked for a data extract via the SAFEHR platform (i.e. a set of parquet files).

As part of the setup prior to this course you were asked to download and install the sample dataset. If you have not done this yet, please refer to the [setup instructions](../learners/setup.md). For now, we will assume that you have the sample OMOP CDM dataset available on your local machine at the following path: `../workshop/data/omop/` and the functions in a folder `../workshop/code/parquet_dataset`.

  
::::::::::::::::::::::::::::::::::::::::::::::::

## Parquet files

Parquet is a columnar storage file format that is optimized for use with big data processing frameworks. It is designed to be efficient in terms of both storage space and read/write performance. Parquet files are often used in data warehousing and big data analytics applications.

## Exploring Parquet files

We have provided a function that will allow you to browse the structure of the data in the same way as we did with the database in the previous episode. This code is available below or in the downloaded `workshop/code/open_omop_dataset.R` file. You can source this file to load the function into your R environment.

```{r, warning = FALSE, message = FALSE}
open_omop_dataset <- function(path) {
    # iterate over table level directories
    list.dirs(path, recursive = FALSE) |>
      # exclude folder name from path and use it as index for named list
      purrr::set_names(~ basename(.)) |>
      # "lazy-load" list of parquet files from specified folder
      purrr::map(arrow::open_dataset)
}
```

**CODING_NOTE**: This function uses the `arrow` package to read in the parquet files. The `open_dataset()` function from the `arrow` package allows us to read in the parquet files without having to load the entire dataset into memory. This is particularly useful when working with large datasets. The function is reasonably complex but it is designed to be flexible and work with any OMOP CDM dataset that is structured in the same way as the one we are using for this course. It will read in all the parquet files in the specified directory and create a nested list structure that allows us to easily access the different tables in the dataset. We leave it to you to explore the code and understand how it works.

Now we can use this function to open the sample OMOP CDM dataset located in the `workshop/data/omop/` directory and explore it in the same way as we did with the database in the previous episode.

```{r, warning = FALSE, message = FALSE}
omop <- open_omop_dataset("./data/omop")
```

::::::::::::::::::::::::::::::::::::: instructor

Note that the path to the data directory may be different depending on where you have stored the sample OMOP CDM dataset on your local machine.

Check the people have used the right path. Their environment should now have an entry under Data reading 'omop List of 8'

::::::::::::::::::::::::::::::::::::::::::::::::

Explore the data using the following:
```{r, warning = FALSE, message = FALSE}
omop
```

You will see that this gives you a list of all the tables in this dataset and what columns they contain. It is obviously a much smaller dataset! You can explore individual tables which will also give you the column names and the data type of the entry.

```{r, warning = FALSE, message = FALSE}
omop$person
```
To actually open each table we can use
```{r, warning = FALSE, message = FALSE}
library(dplyr)
person <- omop$person |>
  collect()

person
```

**CODING_NOTE**: The `collect()` function is used to actually read the data from the parquet files into memory. This is necessary because the `open_dataset()` function creates a reference to the data rather than loading it into memory. By using `collect()`, we can work with the data as a regular data frame in R.

Or we can use the specific functions from the `arrow` package to read in the parquet files directly.

```{r, warning = FALSE, message = FALSE}
library(arrow)
person <- read_parquet("./data/omop/person/part-0.parquet")
person
```

**CODING_NOTE**: The `read_parquet()` function from the `arrow` package allows us to read in a specific parquet file directly into R. This can be useful if we only want to work with a specific table from the dataset and do not need any of the other files.

::::::::::::::::::::::::::::::::::::: instructor

Check that everyone has been able to read in the person table and see the data.

Points worth a discussion:

- the day and month of birth

- gender_source_value and race_source_value are given for some rows but not others

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: callout

As part of the privacy preserving policies around health data, dates of birth are often de-identified to only show the year of birth. This is why in this dataset the day and month of birth are set to 15/6 or 0/0 for all individuals.

You can also see that in some cases the gender_source_value and race_source_value columns are populated while in others they are not. This depends on the policy of the individual hospital. Note: the data in this dataset is a number of different sources combined together to form a single OMOP CDM database.

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: challenge

Adapt the code we had developed for the get_concept_name function in the previous episode to work with this parquet file dataset.

::::::::::::::::::::::::::::::::::::: solution
```{r, warning = FALSE, message = FALSE}
get_concept_name <- function(omop_obj, id) {
  omop_obj$concept |>
    filter(concept_id == id) |>
    select(concept_name) |>
    collect()
}
```

**CODING_NOTE**: The `get_concept_name()` function is adapted to work with the parquet file dataset. It uses the `filter()` and `select()` functions from the `dplyr` package to query the `concept` table in the parquet dataset. The `collect()` function is used to read the result into memory so that we can work with it as a regular data frame in R.

::::::::::::::::::::::::::::::::::::::::::::::::
::::::::::::::::::::::::::::::::::::::::::::::::

Now we can use this function to look up concept names by their concept_id.

```{r, warning = FALSE, message = FALSE}
get_concept_name(omop, 8507)
```

***Answer:*** The concept_id **8507** corresponds to the concept "Male".

::::::::::::::::::::::::::::::::::::: keypoints 

- Parquet files are a columnar storage file format optimized for big data processing.

- The `arrow` package in R can be used to read and manipulate parquet files.

::::::::::::::::::::::::::::::::::::::::::::::::
