---
title: "Dates and times"
teaching: 0
exercises: 0
---

:::::::::::::::::::::::::::::::::::::: questions 

- What is the duration of an event?

- How can I view trends in measurements?:

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Know that dates and times are used in most tables in the OMOP CDM to record when events happened.

- Understand that dates and times are often used in pairs to record the start and end of an event, which allows us to calculate the duration of that event.

- Know how to convert a date or datetime string to a date or datetime object in R.

::::::::::::::::::::::::::::::::::::::::::::::::

## Introduction

This episode considers dates and times in the OMOP Common Data Model (CDM). 

::::::::::::::::::::::::::::::::::::::::::::::::callout

For this episode we will be using a sample OMOP CDM database that is pre-loaded with data. This database is a simplified version of a real-world OMOP CDM database and is intended for educational purposes only.

(UCLH only) This will come in a similar form as you would get data if you asked for a data extract via the SAFEHR platform (i.e. a set of parquet files).

As part of the setup prior to this course you were asked to download and install the sample dataset. If you have not done this yet, please refer to the [setup instructions](../learners/setup.md). For now, we will assume that you have the sample OMOP CDM dataset available on your local machine at the following path: `./data/omop/` and the functions in a folder `./code/parquet_dataset`.

You will then need to load the database as shown in the previous episode.

```{r, warning = FALSE, message = FALSE}
library(dplyr)

open_omop_dataset <- function(path) {
    # iterate over table level directories
    list.dirs(path, recursive = FALSE) |>
      # exclude folder name from path and use it as index for named list
      purrr::set_names(~ basename(.)) |>
      # "lazy-load" list of parquet files from specified folder
      purrr::map(arrow::open_dataset)
}
```

```{r, warning = FALSE, message = FALSE}
omop <- open_omop_dataset("./data/omop")
```

and the useful functions we created in the previous episode to look up concept names and ids.

```{r, warning = FALSE, message = FALSE}
get_concept_name <- function(omop_obj, id) {
  omop_obj$concept |>
    filter(concept_id == id) |>
    select(concept_name) |>
    collect()
}
```

```{r, warning = FALSE, message = FALSE}
get_concept_id <- function(omop_obj, name) {
  omop_obj$concept |>
    filter(concept_name == name) |>
    select(concept_id) |>
    collect()
}
``` 


::::::::::::::::::::::::::::::::::::::::::::::::

# Dates and times in OMOP

Dates and times are used in most tables to record when events happened, usually with the start and end date recorded. In many places there is also a time component, for example to record the time of day when a measurement was taken. Column names are frequently either suffixed with `_date` or `_datetime` to indicate whether they contain just a date or both date and time information.

Parquet has data types for dates and date-times, but if the data was written to CSV or as a string then we can convert these strings to date or datetime objects using the `as.Date()` or `as.POSIXct()` functions, respectively. This will allow us to perform date and time calculations and visualizations more easily.

Usually the dates come in pairs, for example `condition_start_date` and `condition_end_date` in the `condition_occurrence` table. This allows us to determine the duration of a condition, for example, by calculating the difference between the start and end dates.

::::::::::::::::::::::::::::::::::::::::::::::::::: challenge

Using the `condition_occurrence` table, display the duration in days for each condition and find the average duration of conditions in this dataset.

::::::::::::::::::::::::::::::::::::::::::::::::::: solution
```{r, warning = FALSE, message = FALSE}
# First we need to read in the condition_occurrence table with only the columns we need to worry about and collect the data into memory
condition_occurrence <- omop$condition_occurrence |>
  select(condition_occurrence_id, condition_start_date, condition_end_date) |>
    collect()

# Exclude any rows where the end date is missing, as we can't calculate the duration for these
condition_occurrence <- condition_occurrence |>
    filter(!is.na(condition_end_date))

# We can calculate the length of each condition in days by taking the difference between the end date and the start date. We add 1 to include both the start and end date in the duration calculation.
condition_occurrence <- condition_occurrence |>
  mutate(
    length_days_inclusive = as.integer(condition_end_date - condition_start_date) + 1L
  )

# Finally we can calculate the average duration of conditions in this dataset, excluding any missing values
average_duration <- mean(condition_occurrence$length_days_inclusive, na.rm = TRUE)

average_duration
```

***Answer:*** The average duration of conditions in this dataset is approximately 3.2 days.

**CODING NOTE:** In the code above, we first read in the `condition_occurrence` table and selected only the relevant columns. We then filtered out any rows where the `condition_end_date` was missing, as we cannot calculate the duration for these conditions. Next, we calculated the length of each condition in days by taking the difference between the end date and the start date and adding 1 to include both the start and end date in the duration calculation. Finally, we calculated the average duration of conditions in this dataset by taking the mean of the `length_days_inclusive` column, excluding any missing values.
::::::::::::::::::::::::::::::::::::::::::::::::::
::::::::::::::::::::::::::::::::::::::::::::::::

It is not uncommon for the end date to be missing in the data, for example if a condition is ongoing at the time of data extraction. In this case, we can only calculate the duration for conditions that have an end date recorded.

Also consider the `measurement` table where we have both a `measurement_date` and a `measurement_datetime` column. The `measurement_date` column contains just the date of the measurement, while the `measurement_datetime` column contains both the date and time of the measurement. Depending on the analysis we want to do, we may choose to use one or the other of these columns. Note generally a measurement doesn't have a start and end date, but just a single date or datetime when the measurement was taken.

::::::::::::::::::::::::::::::::::::::::::::::::::: challenge

Using the `measurement` table, graph the `Body temperature` of patient `113` over time. You can use either the `measurement_date` or `measurement_datetime` column for the x-axis.

::::::::::::::::::::::::::::::::::::::::::::::::::: solution
```{r, warning = FALSE, message = FALSE}
library(ggplot2)
# First we need to know the concept id for body temperature
get_concept_id(omop, "Body temperature")
```

```{r, warning = FALSE, message = FALSE}

# Then we need to read in the measurement table, filter for the measurements of interest and collect the data into memory
measurement <- omop$measurement |>
    filter(person_id == 1113 & measurement_concept_id == 3020891) |>
    collect() 

ggplot(measurement, aes(x = measurement_datetime, y = value_as_number)) +
  geom_point() +
  geom_line() +
  labs(
    title = "Body temperature of patient 113 over time",
    x = "Date and time of measurement",
    y = "Body temperature (°C)"
  )
```

**CODING NOTE**: In the code above, we first used the `get_concept_id()` function to find the concept ID for "Body temperature". We then read in the `measurement` table and filtered for measurements of patient `113` with the relevant concept ID. We collected this data into memory. Finally, we created a line plot of body temperature over time using `ggplot2`, we did not need to order the data by datetime as ggplot does this for us when we set the datetime column as the value for the x-axis.
::::::::::::::::::::::::::::::::::::::::::::::::::
::::::::::::::::::::::::::::::::::::::::::::::::

## Converting data types

If you are working with a csv file as an input, then there are no guarantees that a dataset will have the expected data types. We can simulate what this may look like by creating a new measurement table that has string types instead of the expected types.

```{r, warning = FALSE, message = FALSE}
measurement_to_fix <- omop$measurement |>
  mutate(
    across(c(measurement_datetime, measurement_date, value_as_number), ~ as.character((.x)))
    ) |>
  collect()

measurement_to_fix
```

We've used the `across()` function from the `dplyr` package to apply the `as.character()` function to three columns.


::::::::::::::::::::::::::::::::::::::::::::::::::: challenge

Using the `measurement_to_fix` table.

1. Calculate the number of days between the first and last `Body temperature` measurement for each patient, sorting the final output by `patient_id`. Which patient has `Body temperature` measurements spanning mulitple days?
2. Create a scatter plot of all `Systolic blood pressure` and `Diastolic blood pressure` measurement values that were taken between the 1st of January 2025 and 1st of October 2025, using `measurement_datetime` for the x-axis. Colour the points by the `patient_id`.

::::::::::::::::::::::::::::::::::::::::::::::::::: solution

1. Calculate the number of days between the first and last `Body temperature` measurement for each patient, sorting the final output by `patient_id`. Which patient has `Body temperature` measurements spanning mulitple days?

```{r, warning = FALSE, message = FALSE}

# Convert the date column to a date
measurement_with_date <- measurement_to_fix |>
    mutate(measurement_date = as.Date(measurement_date, format = "%Y-%m-%d"))

# Calculate the number of days between first and last body temperature measurement
measurement_with_date |>
  filter(measurement_concept_id == 3020891) |>
  summarise(
    first_measurement_date = min(measurement_date),
    last_measurement_date = max(measurement_date),
    days_between = as.integer(last_measurement_date - first_measurement_date),
    .by = person_id
    ) |>
  arrange(person_id)
```
***Answer:*** Only patient 113 has `Body Temperature` measurements over multiple days. 



**CODING NOTE**: In the code above, we converted the `measurement_date` from character to date instances using the `as.Date()` function. We then filtered the data to the concept for Body Temperature, and used `summarise()` grouped by `person_id` to calculate the first and last measurement date, and their difference in numbers of days. 


2. Create a scatter plot of all `Systolic blood pressure` and `Diastolic blood pressure` measurement values that were taken between the 1st of January 2025 and 1st of October 2025, using `measurement_datetime` for the x-axis. Colour the plot by the `patient_id`.


```{r, warning = FALSE, message = FALSE}
# Correct the types of remaining columns
measurement_fixed <- measurement_with_date |>
  mutate(
    measurement_datetime = as.POSIXct(measurement_datetime, tz = "UTC"),
    value_as_number = as.double(value_as_number)
  )

# Get concepts for blood pressure
systolic_concept <- get_concept_id(omop, "Systolic blood pressure") |>
  pull()

diastolic_concept <- get_concept_id(omop, "Diastolic blood pressure") |>
  pull()

# Filter measurements by concepts and time range
measurement_fixed |>
  filter(
    measurement_concept_id %in% c(systolic_concept, diastolic_concept),
    measurement_datetime >= "2025-01-01",
    measurement_datetime < "2025-10-01",
  ) |>
  ggplot(aes(x = measurement_datetime, y = value_as_number, colour = as.factor(person_id))) +
  geom_point() +
  labs(
    title = "Blood pressure measurements over time",
    x = "Measurement taken (datetime)",
    y = "Blood pressure",
    colour = "person_id"
  )

```

**CODING NOTE**: In the code above, we converted the `measurement_datetime` from character to datetime instances using the `as.POSIXct()` function, we would need to confirm which timezone the data is in if we haven't already been told. We also converted the `value_as_number` column to a double (numeric allowing for decimal places) data type using `as.double()`. We then found the concepts required, filtered to these and the date range and used these to plot the data using `ggplot()`. We used the `colour` option in `aes()` to define that the colour of the points should be, converting it to a factor using `as.factor()` so that the continuous variable is diplayed as categories.


::::::::::::::::::::::::::::::::::::::::::::::::::
::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: keypoints 

- Dates and times are used in most tables in the OMOP CDM to record when events happened.

- They are often used in pairs to record the start and end of an event, which allows us to calculate the duration of that event.

::::::::::::::::::::::::::::::::::::::::::::::::::
