---
title: "Chapter 5 Visits and how we can use them"
teaching: 0
exercises: 0
---

:::::::::::::::::::::::::::::::::::::: questions 

- What are visits and how can they be used ?

::::::::::::::::::::::::::::::::::::::::::::::::

::::::::::::::::::::::::::::::::::::: objectives

- Know that a visit is a period of time and patients can have multiple visits
- Understand that multiple measurements, conditions etc. can occur within and between visits
- Understand that for some analyses you will want to look within visits and for other analyses to sum across visits 
- Know that visits are recorded in the visit_occurrence table
- Know each visit is unique to a person
- Understand that other tables link to visits
- Understand how visits can be used to find co-occurrence of other events

::::::::::::::::::::::::::::::::::::::::::::::::

## Introduction

TODO not sure if this wants to be in a separate episode or a more general one about linking tables. It could always be renamed later.

The `visit_occurrence` table contains [events where Persons engage with the healthcare system for a duration of time](https://ohdsi.github.io/CommonDataModel/cdm54.html#visit_occurrence).


The main clinical tables `condition_occurrence`, `measurement`, `observation` and `drug_exposure` contain a `visit_occurrence_id` that links to this table.


`visit_concept_id` specifies the kind of visit that took place using standardised OMOP concepts. These include `Inpatient visit`, `Emergency Room Visit` and `Outpatient Visit`. Inpatient visits can last for longer than one day.    

The `visit_detail` table can contain information about time periods shorter than the visit (for example transfer between wards) but we will not cover that further here.

Firstly here is some code (from ChatGPT) to create some interesting example data.
TODO this code is lengthy, we may later want to hide it or save the output as .Rdata in the repo

```{r, warning = FALSE, message = FALSE}

#person, visit_occurrence, measurement, drug_exposure, condition_occurrence, and concept tables
#realistic blood pressure trends over time
#conditions and drugs co-occurring within visits
#concept table for joining concept names

# Install and load required packages
#install.packages(c("dplyr", "tibble", "lubridate", "uuid"), dependencies = TRUE)

library(dplyr)
library(tibble)
library(lubridate)
library(uuid)
library(tidyr)

set.seed(123)

# 1. Create 100 synthetic patients
n_patients <- 100

person <- tibble(
  person_id = 1:n_patients,
  gender_concept_id = sample(c(8507, 8532), n_patients, replace = TRUE),  # Male / Female
  year_of_birth = sample(1940:2000, n_patients, replace = TRUE)
)

# 2. Create multiple visits per patient
visits_per_person <- sample(2:5, n_patients, replace = TRUE)

visit_occurrence <- tibble(
  person_id = rep(person$person_id, times = visits_per_person)
) |>
  mutate(
    visit_occurrence_id = row_number(),
    visit_start_date = as_date("2020-01-01") + sample(0:1000, n(), replace = TRUE),
    visit_concept_id = sample(c(9201, 9202, 9203), n(), replace = TRUE),  # Inpatient, ER, Outpatient
    #ER & outpatient just a day
    visit_end_date = if_else( visit_concept_id == 9201,
                              visit_start_date + sample(1:3, n(), replace = TRUE),
                              visit_start_date),
  )

# Baseline + slope per patient in one tibble
bp_params <- tibble(
  person_id = 1:n_patients,
  bp_slope = rnorm(n_patients, mean = 0, sd = 0.3),   # mmHg per hour
  systolic_base = rnorm(n_patients, mean = 120, sd = 10)
)

measurement <- visit_occurrence |>
  left_join(bp_params, by = "person_id") |>
  rowwise() |>
  mutate(
    n_hours = as.numeric(interval(visit_start_date, visit_end_date) / hours(1)),
    hours_seq = list(visit_start_date + hours(0:n_hours))
  ) |>
  ungroup() |>
  select(person_id, visit_occurrence_id, bp_slope,
         systolic_base, visit_start_date, hours_seq) |>
  unnest(hours_seq) |>
  mutate(
    hours_since_start = as.numeric(difftime(hours_seq, visit_start_date, units = "hours")),
    systolic = systolic_base +
               (hours_since_start * bp_slope) +   # per-hour slope
               rnorm(n(), 0, 5),                 # noise
    measurement_id = UUIDgenerate(n = n()),
    measurement_concept_id = 3004249,  # systolic
    unit_concept_id = 8510,            # mmHg
    measurement_date = hours_seq
  ) |>
  select(measurement_id, person_id, visit_occurrence_id,
         measurement_date, measurement_concept_id,
         systolic, unit_concept_id) |>
  rename(value_as_number = systolic)


# 6. Define condition-drug co-occurrence map
condition_drug_map <- tribble(
  ~condition_concept_id, ~condition_name,            ~drug_concept_id, ~drug_name,
  201826,                "Type 2 diabetes mellitus",  1124300,          "Metformin",
  320128,                "Essential hypertension",    1112807,          "Lisinopril",
  319835,                "Hyperlipidemia",            19019073,         "Atorvastatin"
)

# 7. Generate conditions per visit
condition_occurrence <- visit_occurrence |>
  rowwise() |>
  do({
    #n_conditions <- sample(1:2, 1)
    #change n_conditions to 1 so that co-occurrence example works
    n_conditions <- 1
    selected_conditions <- condition_drug_map |>
      slice_sample(n = n_conditions)

    tibble(
      condition_occurrence_id = UUIDgenerate(n_conditions),
      person_id = .$person_id,
      visit_occurrence_id = .$visit_occurrence_id,
      condition_concept_id = selected_conditions$condition_concept_id,
      condition_start_date = .$visit_start_date
    )
  }) |>
  bind_rows()

# 8. Generate drug exposures that match conditions
drug_exposure <- condition_occurrence |>
  left_join(condition_drug_map, by = "condition_concept_id") |>
  group_by(person_id, visit_occurrence_id, drug_concept_id) |>
  summarise(
    drug_exposure_id = UUIDgenerate(1),
    drug_exposure_start_date = min(condition_start_date),
    .groups = "drop"
  ) |>
  mutate(days_supply = sample(30:90, n(), replace = TRUE))

# 9. Build concept table (with all used concepts)
concept <- tribble(
  ~concept_id, ~concept_name,
  3004249, "Systolic Blood Pressure",
  3012888, "Diastolic Blood Pressure",
  3027114, "Glucose",
  3016502, "Creatinine",
  1124300, "Metformin",
  1112807, "Lisinopril",
  19019073, "Atorvastatin",
  201826, "Type 2 diabetes mellitus",
  320128, "Essential hypertension",
  319835, "Hyperlipidemia",
  8510, "mm[Hg]",
  8713, "mg/dL",
  8840, "mmol/L",
  9201, "Inpatient Visit",
  9202, "Emergency Room Visit",
  9203, "Outpatient Visit",
  8507, "Male",
  8532, "Female"
)

# 10. Combine into synthetic CDM object
cdm <- list(
  person = person,
  visit_occurrence = visit_occurrence,
  measurement = measurement,
  condition_occurrence = condition_occurrence,
  drug_exposure = drug_exposure,
  concept = concept
)


```


We can query the `visit_occurrence` table to see what kinds of visits there are in the data.

```{r, warning = FALSE, message = FALSE}
library(dplyr)

cdm$visit_occurrence |> 
  count(visit_concept_id) |> 
  left_join(cdm$concept, by = c("visit_concept_id" = "concept_id"))
```

```output
  visit_concept_id     n concept_name        
1             9201   103 Inpatient Visit     
2             9202   112 Emergency Room Visit
3             9203   120 Outpatient Visit    
```

Can you use `person_id` in the `visit_occurrence` table to find patients with more than one visit ?

```{r, warning = FALSE, message = FALSE}
visit_counts <- cdm$visit_occurrence |>
  group_by(person_id) |>
  summarise(n_visits = n()) |>
  filter(n_visits > 1) |>
  collect()

visit_counts |> head(4)
```

```output
  person_id n_visits
1         1        5
2         2        5
3         3        3
4         4        4
```

Now we can choose one of the patients from the previous table and look at all of their visits.

```{r, warning = FALSE, message = FALSE}
example_person_id <- visit_counts$person_id[1]

patient_visits <- cdm$visit_occurrence |>
  filter(person_id == example_person_id) |>
  left_join(cdm$concept, by = c("visit_concept_id" = "concept_id")) |>
  select(
    visit_occurrence_id, visit_start_date, visit_end_date, concept_name
  ) |>
  arrange(visit_start_date) |>
  collect()

patient_visits
```

```output
  visit_occurrence_id visit_start_date visit_end_date concept_name        
1                   4 2020-06-26       2020-06-26     Emergency Room Visit
2                   1 2020-11-30       2020-12-01     Inpatient Visit     
3                   2 2021-04-03       2021-04-04     Inpatient Visit     
4                   5 2021-05-15       2021-05-15     Outpatient Visit    
5                   3 2021-07-12       2021-07-12     Emergency Room Visit
```
Outpatient and Emergency Room Visits usually end on the same day, where Inpatient visits can last longer.

Here is how we can plot a measurement (in this case blood pressure) over time for a patient. The numeric measurement is stored in value_as_number. 

```{r bp-all-visit, warning = FALSE, message = FALSE, fig.alt = "plot of blood pressure measurements for all visits", fig.cap = "plot of blood pressure measurements for all visits"}

# Define concept IDs
systolic_id <- 3004249
selected_patient <- 1

bp <- cdm$measurement |> 
  filter(person_id == selected_patient) |> 
  filter(measurement_concept_id %in% c(systolic_id))


#TODO show the data before plotting

ggplot(bp, aes(x = measurement_date, 
               y = value_as_number)) +
  geom_point() +
  theme_minimal()

```

You should be able to see some dates with single measurements and some with a few measurements very close to each other. Each separate group is a single visit.


::::::::::::::::::::::::::::::::::::: challenge 

Can you modify the query and plot to show the type of visit ?

::::::::::::::::::::::::hint
You can join to visit_occurrence to get visit_concept_id, and from that join to the concept table to get concept_name. You can add `shape =` to the aes() statement in the plot. 
::::::::::::::::::::::::::::

:::::::::::::::::::::::: solution

```{r bp-type-visit, warning = FALSE, message = FALSE, fig.alt = "plot of blood pressure measurements for all visits", fig.cap = "plot of blood pressure measurements for all visits"}

# Define concept IDs
systolic_id <- 3004249
selected_patient <- 1

bp2 <- cdm$measurement |> 
  filter(person_id == selected_patient) |> 
  filter(measurement_concept_id %in% c(systolic_id)) |> 
  left_join(cdm$visit_occurrence, by = c("visit_occurrence_id", "person_id")) |> 
  left_join(cdm$concept, by = join_by(visit_concept_id == concept_id)) |> 
  rename(visit_concept_name = concept_name)

ggplot(bp2, aes(x = measurement_date, y = value_as_number, shape = visit_concept_name)) +
  geom_point() +
  theme_minimal()

```
To indicate the type of visit we need to join to visit_occurrence to get the visit_concept_id and then join to the concept table to get a name for that concept.

Note that we join the visit_occurrence table to the measurement table using both `("visit_occurrence_id", "person_id")`. If we didn't include both one of the columns would get duplicated and renamed.  

:::::::::::::::::::::::::::::::::
:::::::::::::::::::::::::::::::::::::::::::::::


::::::::::::::::::::::::::::::::::::: challenge 

Can you plot the measurements for one of the Inpatient visits ?

::::::::::::::::::::::::hint
You can filter bp created in the previous challenge by one of the visit_occurrence_id from the table earlier. 
::::::::::::::::::::::::::::

:::::::::::::::::::::::: solution

```{r bp-one-visit, warning = FALSE, message = FALSE, fig.alt = "plot of blood pressure measurements for one visit", fig.cap = "plot of blood pressure measurements for one visit"}

bp3 <- bp2 |> filter( visit_occurrence_id == 1 )

ggplot(bp3, aes(x = measurement_date, y = value_as_number, shape = visit_concept_name)) +
  geom_point() +
  theme_minimal()
```

You should be able to see hourly blood pressure measurements within a visit.

:::::::::::::::::::::::::::::::::


:::::::::::::::::::::::::::::::::::::::::::::::



Here is an example where the visit can make a difference to an analysis.
Imagine we want to look at which drugs are associated with which conditions for patients.

If we were to join tables by `person_id` as we have seen in previous exercises we could get conditions and drugs that were separated by many years. Instead we can include `visit_occurrence_id` in the join to get co-occurrence of conditions and drugs in the same visit.

```{r, warning = FALSE, message = FALSE}

co_occurrence_by_visit <- cdm$condition_occurrence |>
  # Join with drug_exposure on person and visit
  inner_join(cdm$drug_exposure, by = c("person_id", "visit_occurrence_id")) |>
  # Count how often each condition–drug pair co-occurs in a visit
  group_by(condition_concept_id, drug_concept_id) |>
  summarise(co_occurrences = n(), .groups = "drop") |>
  
  # Join to get condition name
  left_join(
    cdm$concept |> select(concept_id, concept_name),
    by = c("condition_concept_id" = "concept_id")
  ) |>
  rename(condition_name = concept_name) |>
  
  # Join to get drug name
  left_join(
    cdm$concept |> select(concept_id, concept_name),
    by = c("drug_concept_id" = "concept_id")
  ) |>
  rename(drug_name = concept_name) |>
  select(condition_name, drug_name, co_occurrences)
  
```

Here we see a perfect co-occurrence between conditions & expected drugs for treating that condition (because we generated the example data that way). If instead you didn't use the visit_occurrence_id you would see more unexpected associations occurring across widely separated visits for the same patient.


```output
  condition_name           drug_name    co_occurrences
1 Type 2 diabetes mellitus Metformin               121
2 Hyperlipidemia           Atorvastatin            108
3 Essential hypertension   Lisinopril              106
```

::::::::::::::::::::::::::::::::::::: challenge 

Can you repeat the query without using `visit_occurrence_id` to see what results you get ?


```{r, warning = FALSE, message = FALSE}
co_occurrence <- cdm$condition_occurrence |>
  # Join with drug_exposure on person
  inner_join(cdm$drug_exposure, by = c("person_id")) |>
  # Count how often each condition–drug pair co-occurs
  group_by(condition_concept_id, drug_concept_id) |>
  summarise(co_occurrences = n(), .groups = "drop") |>
  
  # Join to get condition name
  left_join(
    cdm$concept |> select(concept_id, concept_name),
    by = c("condition_concept_id" = "concept_id")
  ) |>
  rename(condition_name = concept_name) |>
  
  # Join to get drug name
  left_join(
    cdm$concept |> select(concept_id, concept_name),
    by = c("drug_concept_id" = "concept_id")
  ) |>
  rename(drug_name = concept_name) |>
  select(condition_name, drug_name, co_occurrences)
```

Now we see that conditions co-occur with drugs unlikely to be used in their treatment because they came from visits further apart in time.

```output
  condition_name           drug_name    co_occurrences
1 Type 2 diabetes mellitus Lisinopril               99
2 Type 2 diabetes mellitus Metformin               251
3 Type 2 diabetes mellitus Atorvastatin            112
4 Hyperlipidemia           Lisinopril               86
5 Hyperlipidemia           Metformin               112
6 Hyperlipidemia           Atorvastatin            222
7 Essential hypertension   Lisinopril              190
8 Essential hypertension   Metformin                99
9 Essential hypertension   Atorvastatin             86
```

:::::::::::::::::::::::::::::::::::::::::::::::


::::::::::::::::::::::::::::::::::::: keypoints 



::::::::::::::::::::::::::::::::::::::::::::::::

[r-markdown]: https://rmarkdown.rstudio.com/
